---
title: "Basic qQTL Analysis with qQTLR: Calling qQTLs and Assessing Heterogeneity"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic qQTL Analysis with qQTLR: Calling qQTLs and Assessing Heterogeneity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Quantile QTL (qQTL) analysis models how genetic variants influence the
**entire distribution** of a molecular phenotype rather than its mean alone.
Compared with linear xQTL, this captures **non-linear**, **context-dependent**
regulatory effects — including variants that act primarily at the
distributional extremes or whose effect sizes vary across expression levels.

This vignette walks through a minimal qQTL workflow with `qQTLR`:

1. Screen variants with the **QRank** score test across a grid of quantiles.
2. Combine quantile-specific p-values across the distribution.
3. Assess **distributional heterogeneity** via the Chatterjee correlation test
   (implemented in `calculate_xi_correlation()`).
4. Classify variants as *qQTL-only*, *shared-homogeneous*, or
   *shared-heterogeneous* relative to a linear xQTL reference.

## Setup

```{r load-pkgs}
library(qQTLR)
```

## Step 1 — Screen quantile associations

`qr_screen()` performs a fast QRank-based screen across a grid of quantiles
(τ = 0.1, 0.15, …, 0.9 by default) and returns per-variant p-values at each
quantile, plus a Cauchy-combined p-value across quantiles.

```{r qr-screen}
# Placeholder — replace X, y, covariates with your data
# qr_screen_result <- qr_screen(
#   X    = genotype_matrix,
#   y    = phenotype_vector,
#   Z    = covariate_matrix,
#   tau  = seq(0.1, 0.9, by = 0.05)
# )
```

## Step 2 — Fit full quantile regressions at significant variants

Follow up the screen with full quantile regressions at the selected variants
to obtain quantile-specific effect sizes and standard errors:

```{r qr-fit}
# perform_qr_analysis(
#   X   = genotype_matrix[, top_variants],
#   y   = phenotype_vector,
#   Z   = covariate_matrix,
#   tau = seq(0.1, 0.9, by = 0.05)
# )
```

## Step 3 — Assess distributional heterogeneity

Among variants that are also significant in linear xQTL ("shared" qQTLs), use
the Chatterjee ξ correlation to test whether their effect sizes vary across
quantiles. A significant ξ indicates a *shared-heterogeneous* qQTL — the mean
effect is real, but the underlying regulatory architecture is distributional.

```{r xi}
# calculate_xi_correlation(tau_grid, beta_tau)
```

## Step 4 — Classify variants

Collectively, **qQTL-het** = *qQTL-only* ∪ *shared-heterogeneous*; these are
the variants that carry information beyond the linear model and are the main
biological payload of a qQTL scan.

| Category                    | Linear xQTL significant? | qQTL significant? | Heterogeneity (ξ) |
|-----------------------------|:------------------------:|:-----------------:|:-----------------:|
| `xQTL-only`                 | ✓                        | ✗                 | —                 |
| `shared-homogeneous`        | ✓                        | ✓                 | not significant   |
| `shared-heterogeneous`      | ✓                        | ✓                 | significant       |
| `qQTL-only`                 | ✗                        | ✓                 | —                 |

## Next steps

- [Interpreting qQTL analysis results](interpreting-qqtl-results.html)
- [vQTL as a special case of qQTL](vqtl-special-case.html)
- [The qQTL TWAS pipeline](qtwas-pipeline.html)
