---
title: "Interpreting qQTL Analysis Results"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpreting qQTL Analysis Results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## From significance to biology

Quantile-specific p-values and effect sizes are only the starting point.
The biological value of a qQTL scan comes from reading the **shape** of the
effect across the quantile grid (τ = 0.1, …, 0.9). This vignette outlines
recurring patterns and how they map onto regulatory mechanisms, drawing on
examples from the qQTLR manuscript.

## Canonical quantile-effect patterns

```{r load}
library(qQTLR)
```

### 1. Symmetric tail effects → variance QTL (vQTL)

Opposite-sign effects at τ = 0.1 and τ = 0.9 with ≈ 0 effect at the median
indicate a *variance-increasing* genotype that does not shift the mean.
Linear regression is blind to these variants; vQTL methods such as QUAIL
can recover them and qQTL captures them as a special case.

*Example from the manuscript:* `PITRM1` (chr10:3147188:C>T) in ROSMAP AC —
linear β = 0.108, p = 0.0126; vQTL p = 3.97 × 10⁻¹⁴; qQTL p = 1.72 × 10⁻⁷.

### 2. One-tail enrichment → context-specific effects at phenotypic extremes

Significance only at τ = 0.1 *or* τ = 0.9 suggests the variant acts only in
a vulnerable sub-population (e.g., individuals with already-low expression,
or a specific cell-type composition, or a specific interaction background).

*Example from the manuscript:* `TMEM106B` (chr7:12086129:A>G) in excitatory
neurons — significant at τ = 0.1–0.2 only; follow-up iQTL testing revealed
the pattern is partly explained by *APOE* ε4 interaction.

### 3. Monotone effect across quantiles → progressive regulatory gradient

Effect size that monotonically strengthens from low to high (or vice versa)
quantiles suggests coordinated mechanisms that compound as baseline
expression shifts — often involving multiple epigenetic layers.

*Example from the manuscript:* `CHRNE` vs. `SCIMP` at chr17 — opposite
directions of the quantile gradient, linked to coordinated DNA methylation,
H3K9ac and chromatin accessibility changes.

## Deciding which follow-ups are warranted

| Observed pattern                          | Suggested follow-up                                         |
|-------------------------------------------|-------------------------------------------------------------|
| Symmetric tails, no mean effect           | Confirm with vQTL (QUAIL); look for latent interactions     |
| One-tail only, plausible interacting var  | Targeted iQTL test against candidate factors (sex, APOE…)   |
| Monotone gradient                         | Epigenomic colocalization (caQTL / mQTL / haQTL); mediation |
| qQTL-only + enriched annotations          | Cross-cell-type regulatory element / chromatin loop lookup  |

## Heterogeneity vs. discovery

`qQTL-het` — the union of qQTL-only and shared-heterogeneous variants —
represent regulatory effects that **cannot** be recovered by linear
modeling alone. The manuscript reports ≈ 28% of qQTLs are qQTL-het across
contexts, with the proportion varying up to ≈ 50% in some pQTL contexts.

## Caveats

- Quantile regression is *less* powerful than linear regression for pure
  mean effects. Do not replace xQTL with qQTL — use them together.
- Tail quantiles depend on sample size and distributional tails; extreme
  effects at τ = 0.05 or 0.95 can be unstable in small cohorts.
- Always sanity-check tail effects against raw genotype-stratified
  distributions before reporting a "variance QTL" or "extreme-only" effect.

## See also

- [Basic qQTL analysis](basic-qqtl-analysis.html)
- [vQTL as a special case of qQTL](vqtl-special-case.html)
- [iQTL simulation settings](iqtl-simulation.html)
