---
title: "Simulation Settings for iQTL and Detection via qQTL"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Settings for iQTL and Detection via qQTL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Motivation

Interaction QTLs (iQTLs) test for context-dependent regulation (sex,
age, *APOE* ε4 dosage, cell-type fraction, …) using a parametric
G + E + G×E model. They require the interacting factor to be specified
in advance and — as shown in the qQTLR manuscript — have limited power
and robustness at realistic sample sizes.

qQTL can be viewed as a **factor-agnostic screen** for some classes of
interaction effects: when the interaction causes the genotype to act
differently across expression strata, the signal shows up as
distributional heterogeneity.

This vignette describes the simulation settings used in the manuscript
to compare iQTL, qQTL and linear xQTL, and how to reproduce them.

## Interaction taxonomy

Following Kasela et al. (2024), stratified genetic effects are
categorized by the sign of the main and interaction effects:

| Category        | Main effect          | Interaction effect  | Linear xQTL visible? | qQTL visible?     |
|-----------------|----------------------|---------------------|:--------------------:|:-----------------:|
| *Amplifying*    | same sign as G×E     | same sign           | ✓ (strong)           | ✓ (often het.)    |
| *Counteracting* | opposite sign to G×E | opposite sign       | often cancels        | ✓ (better power)  |
| *Uncertain*     | ≈ 0                  | significant         | ✗                    | weak              |

The empirical observation from the manuscript: qQTL recovers ≈ 99% of
amplifying iQTLs, but ≤ 20% of counteracting and ≤ 35% of uncertain
iQTLs.

## Simulation recipe

Parameters calibrated to FunGen-xQTL-scale data:

- Sample size: n ≈ 300 – 1000 per context.
- MAF: U(0.05, 0.5); additive coding 0/1/2.
- Interacting factor E: binary or continuous (e.g., cell fraction).
- Main effect β_G, interaction effect β_{G×E} chosen to produce each
  regime (amplifying / counteracting / uncertain) with matched total
  variance explained.
- Error distribution: Gaussian (primary) and heavy-tailed (sensitivity).

For each simulated phenotype, analyze with:

1. Linear xQTL (`lm(y ~ G + E + covariates)`).
2. iQTL (`lm(y ~ G * E + covariates)`; test G×E term).
3. qQTL (`qr_screen` across τ = 0.1 – 0.9).

```{r load}
library(qQTLR)
# Placeholder — a self-contained simulation helper for interaction regimes
# will be added in a future release. For now, see the manuscript Methods.
```

## Key results to reproduce

- **Well-powered regime (α = 0.05):** amplifying iQTLs ≈ 99% overlap
  between qQTL and xQTL; counteracting iQTLs preferentially detected by
  qQTL; uncertain iQTLs poorly detected by either.
- **Genome-wide regime (α = 1 × 10⁻⁶):** detection rates drop sharply
  for all methods. qQTL retains a modest edge for counteracting
  interactions but absolute power is low.
- Among amplifying iQTLs that qQTL picks up at α = 0.05, ≈ 95% exhibit
  Chatterjee ξ heterogeneity (p_ξ < 0.05).

## Practical takeaway

qQTL is a useful **screening layer** to prioritize loci for targeted
iQTL follow-up with specific candidate factors (see the `TMEM106B`
*APOE* ε4 example). It does **not** replace comprehensive iQTL testing:
counteracting and uncertain interactions without distributional
signatures will be missed.

## See also

- [Basic qQTL analysis](basic-qqtl-analysis.html)
- [vQTL as a special case of qQTL](vqtl-special-case.html)
- [Interpreting qQTL results](interpreting-qqtl-results.html)
