---
title: "The qQTL TWAS (qTWAS) Pipeline"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The qQTL TWAS (qTWAS) Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Quantile-based transcriptome-wide association study (**qTWAS**) extends
standard TWAS by training prediction models at multiple quantiles of the
molecular phenotype, then combining the resulting gene–trait tests across
quantiles. The goal is to recover gene associations with GWAS traits
whose pathogenic effects are **distributional** — e.g., threshold-
dependent immune dysfunction or telomere-length regulation — and are
missed by mean-based TWAS.

In the manuscript, qTWAS recovers 34 novel AD gene associations and 90
qTWAS-only genes for telomere length across 18 brain molecular contexts.

## Pipeline entry points in qQTLR

- `quantile_twas_weight_pipeline()` — train per-gene, per-context qTWAS
  prediction weights at a grid of quantiles.
- `quantile_twas_pipeline()` — apply trained weights to GWAS summary
  statistics and produce gene-level z-scores / p-values per quantile,
  combined across quantiles (Cauchy combination).
- `load_quantile_twas_weights()` — load pretrained qTWAS weights from the
  published brain resource.
- `multicontext_ld_clumping()` — clump across contexts to resolve
  correlated gene-context hits. (Delegates to `pecotmr`'s LD clumping /
  pruning internals.)

## Step 1 — Build qTWAS weights

```{r load}
library(qQTLR)
```

```{r weights}
# quantile_twas_weight_pipeline(
#   genotype_file = "genotypes.bed",
#   phenotype     = expression_matrix,
#   covariates    = covariate_matrix,
#   region        = "cis-window or TADB",
#   tau           = seq(0.05, 0.95, by = 0.05),
#   output        = "weights/"
# )
```

Each gene × context × τ combination yields a vector of SNP weights. The
manuscript trains ~70 000 such gene-context models across 18 brain
contexts.

## Step 2 — Apply weights to GWAS summary statistics

```{r twas}
# twas_result <- quantile_twas_pipeline(
#   weights         = loaded_weights,
#   gwas_sumstats   = gwas_df,
#   ld_reference    = ld_ref,
#   combine_method  = "cauchy"
# )
```

Internally, per-quantile z-scores are built with
`pecotmr::build_twas_score_row()` and combined across quantiles. The
output contains:

- Per-quantile TWAS z/p;
- Cauchy-combined p-value across quantiles;
- A flag for whether the gene is **qTWAS-only** relative to a standard
  TWAS reference.

## Step 3 — Multi-context LD clumping

When a gene is tested across many molecular contexts (bulk tissues,
cell types, proteins), signals correlate. Use
`multicontext_ld_clumping()` to resolve these to a parsimonious set of
independent hits before downstream interpretation.

```{r clump}
# clumped <- multicontext_ld_clumping(twas_result, ld_reference = ld_ref)
```

## Using the pre-trained brain qTWAS resource

The manuscript publishes qTWAS weights across 18 FunGen-xQTL brain
contexts (bulk-tissue eQTL across 9 regions, single-nucleus eQTL across
6 cell types, pQTL across 3 regions). Load them with
`load_quantile_twas_weights()` and apply them to any external GWAS.

```{r resource}
# weights <- load_quantile_twas_weights("path/or/synapse/id")
```

## Interpreting qTWAS-only genes

- Pathway enrichment in the manuscript points to **immune regulation**,
  **lipid metabolism**, and **tau binding** for AD qTWAS-only genes, and
  to **telomere maintenance** machinery for telomere-length qTWAS-only
  genes.
- qTWAS-only hits with suggestive (p < 1 × 10⁻⁴) linear TWAS signals are
  generally the most robust; isolated qTWAS signals without any
  mean-level support warrant extra scrutiny.

## See also

- [Basic qQTL analysis](basic-qqtl-analysis.html)
- [Interpreting qQTL results](interpreting-qqtl-results.html)
- [vQTL as a special case of qQTL](vqtl-special-case.html)
- [iQTL simulation settings](iqtl-simulation.html)
