---
title: "Training qTWAS Prediction Weights"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Training qTWAS Prediction Weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Scope

This vignette covers **training quantile-based TWAS (qTWAS) prediction
weights** with `qQTLR`. Weight *application* to GWAS summary statistics
— the actual gene–trait association test — is not performed inside
`qQTLR`; it is run through the FunGen-xQTL protocol pipelines (see
[Notes on the qQTL manuscript pipelines](manuscript-pipeline-notes.html)).

## Background

qTWAS extends standard TWAS by training prediction models across a grid
of quantiles of the molecular phenotype rather than just its mean.
Combining these models downstream captures gene–trait associations
driven by distributional effects that mean-based TWAS misses.

## Pipeline entry points in qQTLR

- `quantile_twas_weight_pipeline()` — train per-gene, per-context qTWAS
  prediction weights across a grid of quantiles.
- `quantile_twas_pipeline()` — helper that orchestrates the
  weight-training workflow for a full cis-region.
- `load_quantile_twas_weights()` — load previously trained weight
  objects from disk.
- `multicontext_ld_clumping()` — LD-based clumping / pruning across
  contexts (delegates to `pecotmr`).

## Train weights for a single gene × context

```{r load}
library(qQTLR)
```

```{r weights}
# quantile_twas_weight_pipeline(
#   genotype_file = "genotypes.bed",
#   phenotype     = expression_matrix,
#   covariates    = covariate_matrix,
#   region        = "cis-window or TADB",
#   tau           = seq(0.05, 0.95, by = 0.05),
#   output        = "weights/"
# )
```

Each gene × context × τ combination yields a vector of SNP weights.

## Multi-context LD clumping

When training across many molecular contexts, correlated signals can be
resolved to a parsimonious set using `multicontext_ld_clumping()`:

```{r clump}
# clumped_weights <- multicontext_ld_clumping(weight_set, ld_reference = ld_ref)
```

## Loading pretrained brain qTWAS weights

The qQTLR manuscript publishes trained weights across 18 FunGen-xQTL
brain contexts. Load them with:

```{r resource}
# weights <- load_quantile_twas_weights("path/or/synapse/id")
```

## Running the weights against a GWAS

Weight application — cTWAS and related association testing — is handled
by the FunGen-xQTL protocol pipeline, not inside `qQTLR`. See the
[manuscript pipelines vignette](manuscript-pipeline-notes.html) for the
relevant workflow link.

## See also

- [Basic qQTL analysis](basic-qqtl-analysis.html)
- [Interpreting qQTL results](interpreting-qqtl-results.html)
- [Notes on the qQTL manuscript pipelines](manuscript-pipeline-notes.html)
