---
title: "vQTL as a Special Case of qQTL"
author: "qQTLR developers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vQTL as a Special Case of qQTL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Motivation

Variance QTLs (vQTLs) identify variants that change the *spread* of a
molecular phenotype. `qQTLR` implements the **QUAIL** rank-score approach
for vQTL discovery, and — more importantly — shows empirically and
theoretically that **vQTL is a special case of qQTL**: a variant with
symmetric, opposite-sign effects at the lower and upper quantiles
(around τ ≈ 0.1 and τ ≈ 0.9) with no median effect will be picked up by
qQTL as well as by vQTL.

In the qQTLR manuscript, 93.66% of significant vQTL associations were
recovered by qQTL, predominantly at both tail quantiles simultaneously,
confirming the nested relationship.

## Running QUAIL-based vQTL analysis

```{r load}
library(qQTLR)
```

`qQTLR` exposes two pipeline entry points for vQTL analysis:

- `QUAIL_rank_score_pipeline()` — compute the QUAIL rank score from a
  phenotype and its covariates across a grid of quantiles.
- `QUAIL_pipeline()` — run the end-to-end vQTL test that regresses the
  rank score on genotypes.

```{r quail}
# quail_scores <- QUAIL_rank_score_pipeline(
#   y   = phenotype_vector,
#   Z   = covariate_matrix,
#   tau = seq(0.05, 0.95, by = 0.05)
# )
#
# vqtl_result <- QUAIL_pipeline(
#   X      = genotype_matrix,
#   scores = quail_scores,
#   Z      = covariate_matrix
# )
```

## Reconciling vQTL with qQTL output

For the same variant, compare:

- the **vQTL p-value** from `QUAIL_pipeline()`;
- the **qQTL Cauchy-combined p-value** from `qr_screen()`;
- the **per-quantile effect sizes** from `perform_qr_analysis()`.

Expected joint pattern for a pure variance effect:

- vQTL significant.
- qQTL significant with comparable p-value.
- Effect sizes symmetric in magnitude and opposite in sign at the low and
  high tails, ≈ 0 at the median.
- Linear xQTL may be non-significant.

Asymmetric distributional effects — for example lower-tail-only — will
appear as qQTL signals but may be missed by vQTL, which assumes symmetric
spread. This is why qQTL is strictly more general.

## When to use which

| Situation                                               | Recommended tool     |
|---------------------------------------------------------|----------------------|
| Symmetric variance changes only                         | QUAIL vQTL           |
| Full distributional scan (tails, asymmetry, gradients)  | qQTL via `qr_screen` |
| Screening → mechanism follow-up                         | qQTL then targeted   |

## See also

- [Basic qQTL analysis](basic-qqtl-analysis.html)
- [Interpreting qQTL results](interpreting-qqtl-results.html)
- [iQTL simulation settings](iqtl-simulation.html)
